This thesis studies task-driven LiDAR compression for autonomous driving under fixed downstream evaluation. The central thesis is that compression should be judged not only by generic signal fidelity, but also by whether it preserves the information needed by downstream perception models. To make that question concrete, the work is organized into two evidence lanes. Track~2 is the main thesis path: an in-repository range-image compression pipeline evaluated under a frozen RangeDet compare bundle on KITTI. Track~3 is a bounded support path built on an external point-cloud-native codec (RENO) with a fixed PointPillars endpoint. This separation allows the thesis to present one main claim and one supporting baseline family without conflating them.

The current draft reports completed results for both lanes. On Track~2, the confirmed Stage~1 result reached \(AP3D@0.3=0.2492\), \(AP3D@0.5=0.1266\), and \(meanBestIoU3D@0.3=0.3170\). A bounded Stage~2 adaptive follow-up improved those values further to \(AP3D@0.3=0.2717\), \(AP3D@0.5=0.1539\), and \(meanBestIoU3D@0.3=0.3611\), while also lowering the internal rate proxy relative to the confirmed Stage~1 baseline. On Track~3, bounded \(B0\) and \(B1\) operating points at \(posQ\in\{16,64\}\) completed with paired PointPillars endpoints, and the resulting baseline references were stable: \(B_{ref,16}=36.4479\), \(B_{ref,64}=36.0382\), with drift below \(0.5\%\) at both operating points. These findings support two conclusions. First, task-driven adaptive compression is a viable and measurable direction when the evaluator bundle is frozen and auditable. Second, an external support baseline can preserve substantial downstream utility across operating points, but it should remain support evidence rather than replace the main claim path. The remaining work is to finalize the bounded Track~3 adaptive attempt and to package the current evidence into publication-quality figures.
