\section{Artifact Map and Figure Plan}\label{apx:artifacts}

\subsection{Canonical Result Groups}
The thesis currently relies on three canonical result groups:

\begin{itemize}
\item \textbf{Track~2 mainline results}: the confirmed Stage~1 result and the bounded Stage~2 adaptive A0 result under the frozen RangeDet compare bundle.
\item \textbf{Track~3 support baselines}: bounded \(B0\) and \(B1\) operating points at \(posQ=16\) and \(posQ=64\) with paired PointPillars endpoints.
\item \textbf{Track~3 adaptive attempt}: a bounded dual-\(posQ\) adaptive follow-up that is still in progress and is therefore not yet treated as final evidence in the main body.
\end{itemize}

\subsection{Figures Missing From the Current Draft}
The current draft now includes the following main-body figures:

\begin{enumerate}
\item \textbf{Thesis system overview figure}. A clean diagram showing Track~2 as the main claim lane and Track~3 as the support lane, with the frozen downstream evaluators highlighted.
\item \textbf{Track~2 progression figure}. A compact plot showing Stage~0 winner, Stage~1 recovered, Stage~1 confirm, and Stage~2 adaptive A0 on the same detector-facing axis.
\item \textbf{Track~3 operating-point figure}. A plot of \(true\ bpp\) versus \(map3d\_mod\_mean\) for \(B0\) and \(B1\) at \(posQ=16\) and \(posQ=64\).
\item \textbf{Track~3 drift figure}. A small paired comparison showing why the adaptive gate remains open after \(B_{ref}\) is defined.
\end{enumerate}

\subsection{Tables Already Included in the Main Body}
The current draft also includes two thesis-facing tables in the main body:
\begin{itemize}
\item \textbf{Track~2 mainline result table}, covering the Stage~0 winner, Stage~1 recovered run, Stage~1 confirm run, and Stage~2 adaptive A0 result.
\item \textbf{Track~3 support baseline table}, covering \(B0@16\), \(B0@64\), \(B1@16\), and \(B1@64\) with downstream metrics and bitrate context.
\end{itemize}

\subsection{Optional Stretch Figures}
If time permits, the next revision should also include:
\begin{itemize}
\item a qualitative reconstruction panel for representative LiDAR frames, and
\item one figure for the bounded Track~3 adaptive attempt, but only after that experiment is canonical.
\end{itemize}

\subsection{Draft Honesty Policy}
This revision intentionally prioritizes the required draft sections---abstract, introduction, related work, and method---plus the strongest currently frozen results. Any unfinished adaptive lane should remain explicitly labeled as ongoing validation rather than being silently blended into the final claim.
